Reproducible Research 
========================================================

# Peer Assessment 1 Assignment

## Loading and processing the data

Here we assume that **activity.csv** is in the current working directory, then use *read.csv()* to read 
data from this file, and storing into *activity* variable.

Since reading is usually a resource intensive task, this operation is being cached.

```{r reading, cache=TRUE}
activity <- read.csv(file="activity.csv", header=TRUE)
```

Then, let's check the first rows of data.

*Note: Assume xtable is already installed.*

```{r reading.check, results="asis"}
# using results="asis"
library(xtable)
xt <- xtable(head(activity))
print(xt, type="html", NA.string="<font color=\"red\">NA</font>")
```

Before proceeding, let's check the structure of *activity*:

```{r check.structure}
str(activity)
```

We need to adjust the date column from Factor to a more appropriate type:
```{r transform.factor2date}
activity <- transform(activity, date = strptime(date, format="%Y-%m-%d"))
str(activity)
```

Now it looks okay:

*Note: xtable misinterprets date class, let's use plain R formatting*
```{r reading.check2}
print(head(activity))
```


## What is mean total number of steps taken per day?

First, let's remove *NA* values:

```{r filter.NA}
activity.filter <- activity[!(is.na(activity$steps)),]
dim(activity)
dim(activity.filter)
```

Let's make a histogram of the total number of steps taken each day:

```{r hist.steps, fig.width=10}
plot(activity.filter$date, activity.filter$steps, type="h", xlab="date", ylab="steps")
```

Now let's calculate the mean and median total number of steps per day

```{r mean.median, results='asis'}
# using results="asis"
steps.means <- aggregate(steps ~ as.character(date), data=activity.filter, FUN=mean)
names(steps.means) <- c("date", "steps")

steps.medians <- aggregate(steps ~ as.character(date), data=activity.filter, FUN=median)
names(steps.medians) <- c("date", "steps")

xt <- xtable(steps.means)
print(xt, type="html")
xt <- xtable(steps.medians)
print(xt, type="html")
```

## What is the average daily activity pattern?


Let's make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)
and the average number of steps taken, averaged across all days (y-axis), then let's identify
which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps.

```{r steps.by.interval}
steps.means.interval <- aggregate(steps ~ interval, data=activity, FUN=mean)
plot(x=steps.means.interval$interval, y=steps.means.interval$steps, type="l", xlab="interval", ylab="steps")

steps.means.interval.peak <- steps.means.interval[steps.means.interval$steps == max(steps.means.interval$steps),]
with(steps.means.interval.peak, points(interval, steps, pch=8, col="red"))

print(steps.means.interval.peak)
```

The point marked in red above represents the highest average of **`r steps.means.interval.peak$steps`** steps
at the **`r steps.means.interval.peak$interval`** interval (hhmm).

## Imputing missing values

Let's calculate and report the total number of missing values in the dataset:

```{r report.nas}
table(is.na(activity$steps))
```

Then let's fill these *NA*s with the average value of steps for that interval, in a copy of our dataset:
```{r fill.nas}
activity.new <- activity
for ( i in 1:nrow(activity) ) {
    if ( is.na(activity[i,]$steps) ) {
        steps <- steps.means.interval[steps.means.interval$interval == activity[i,]$interval,]$steps
        activity.new[i,]$steps <- steps
    }
}
table(is.na(activity$steps))
```

Let's compare our previous histogram with one with the filled data

```{r fill.nas.histogram, fig.width=15}
par(mfrow=c(1,2))
plot(activity.filter$date, activity.filter$steps, type="h", main="original", xlab="date", ylab="steps")
plot(activity.new$date, activity.new$steps, type="h", main="filled", xlab="date", ylab="steps")
```

Now let's calculate the mean and median total number of steps per day with the imputed data

```{r mean.median.compare, results='asis'}
# using results="asis"
steps.means.new <- aggregate(steps ~ as.character(date), data=activity.new, FUN=mean)
names(steps.means.new) <- c("date", "steps")

steps.medians.new <- aggregate(steps ~ as.character(date), data=activity.new, FUN=median)
names(steps.medians.new) <- c("date", "steps")

steps.means.compare <- merge(steps.means, steps.means.new, by="date", suffixes=c(".original",".filled"))
steps.medians.compare <- merge(steps.medians, steps.medians.new, by="date", suffixes=c(".original",".filled"))

xt <- xtable(steps.means.compare)
print(xt, type="html")
xt <- xtable(steps.medians.compare)
print(xt, type="html")
```

Let's check:

1. Do these values differ from the estimates from the first part of the assignment?

2. What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r check.compare}
with(steps.means.compare, table(steps.original - steps.filled))
with(steps.medians.compare, table(steps.original - steps.filled))
```

It's possible to see that, though the histogram shows the extra values for the imputed data, 
the average and medians per day are not affected by the imputed values


## Are there differences in activity patterns between weekdays and weekends?

*Note: For this step I need to set the locale, because my environment is not in english*
```{r locale}
Sys.setlocale(category = "LC_ALL", locale = "C")
```

Let's create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r create.factors}
wday.type <- function(wday) {
    if (wday == "Sat" | wday == "Sun") {
        return("weekend")
    } else {
        return("weekday")
    }
}

wday <- weekdays(strptime(activity.new$date,format="%Y-%m-%d"), abbreviate=T)

weektype <- sapply(wday, wday.type)

activity.new$weektype <- as.factor(weektype)
```

Then, let's make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

*Note: assume ggplot2 package is installed*

```{r weekend.weekdays.plot, fig.width=10}
library(ggplot2)
mean.interval.weektype <- aggregate(steps ~ interval + weektype, data=activity.new, FUN=mean)
qplot(x=interval, y=steps, data=mean.interval.weektype, facets=weektype~., geom="line")
```
